---
title: "CRISIS_Factor_Pipeline EFA co oc problem behaviors 3/25 Split 1"
author: "Aki Nikolaidis, Paul Bloom, Michelle VanTieghem"
date: "3/25/21"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---
Edited Factor analysis from CRISIS General - 8/11/20 BV & JD



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(viridis)
library(readxl)
library(corrplot)
library(knitr)
library(REdaS) # for diagnostic data checks before FA
library(psych)# need for factor analysis
library(GPArotation) # need for factor analysis
library(polycor)
library(lavaan)
library(sjmisc)

```

#DATA ORGANIZATION
```{r, Data_Organization}
All_Site_Split1 <- read.csv("03_25_Split_1_across_All_sites.csv", header = TRUE, sep = ",")

#Converting all ratings to severity score by problem behavior   
All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbHyper_3M = ifelse(hyperactivity == "0", "0",  
                               ifelse(hyperactivity == "1" & hyperactivityhowmuch == "1", "1",  
                                      ifelse(hyperactivity == "1" & hyperactivityhowmuch == "2","2",  
                                             ifelse(hyperactivity == "1" & hyperactivityhowmuch == "3","3",  
                                                    ifelse(hyperactivity == "1" & hyperactivityhowmuch == "4","4", "check"))))))  

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbOnTask_3M = ifelse(stayingontask == "0", "0",  
                                ifelse(stayingontask == "1" & stayingontaskhowmuch == "1", "1",  
                                       ifelse(stayingontask == "1" & stayingontaskhowmuch == "2","2",  
                                              ifelse(stayingontask == "1" & stayingontaskhowmuch == "3","3",  
                                                     ifelse(stayingontask == "1" & stayingontaskhowmuch == "4","4", "check"))))))  


All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbAngry_3M = ifelse(angrytemper == "0", "0",  
                               ifelse(angrytemper == "1" & angrytemperhowmuch == "1", "1",  
                                      ifelse(angrytemper == "1" & angrytemperhowmuch == "2","2",  
                                             ifelse(angrytemper == "1" & angrytemperhowmuch == "3","3",  
                                                    ifelse(angrytemper == "1" & angrytemperhowmuch == "4","4", "check"))))))      

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbVerbAgg_3M = ifelse(verbalaggression == "0", "0",  
                                 ifelse(verbalaggression == "1" & verbalaggressionhowmuch == "1", "1",  
                                        ifelse(verbalaggression == "1" & verbalaggressionhowmuch == "2","2",  
                                               ifelse(verbalaggression == "1" & verbalaggressionhowmuch == "3","3",  
                                                      ifelse(verbalaggression == "1" & verbalaggressionhowmuch == "4","4", "check"))))))      


All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbPhysAgg_3M = ifelse(physicalaggression == "0", "0",  
                                 ifelse(physicalaggression == "1" & physicalaggressionhowmuch == "1", "1",  
                                        ifelse(physicalaggression == "1" & physicalaggressionhowmuch == "2","2",  
                                               ifelse(physicalaggression == "1" & physicalaggressionhowmuch == "3","3",  
                                                      ifelse(physicalaggression == "1" & physicalaggressionhowmuch == "4","4", "check"))))))      

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbDibInjS_3M = ifelse(injureself == "0", "0",  
                                 ifelse(injureself == "1" & injureselfhowmuch == "1", "1",  
                                        ifelse(injureself == "1" & injureselfhowmuch == "2","2",  
                                               ifelse(injureself == "1" & injureselfhowmuch == "3","3",  
                                                      ifelse(injureself == "1" & injureselfhowmuch == "4","4", "check"))))))      

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbDisobArg_3M = ifelse(disobedient == "0", "0",  
                                  ifelse(disobedient == "1" & disobedienthowmuch == "1", "1",  
                                         ifelse(disobedient == "1" & disobedienthowmuch == "2","2",  
                                                ifelse(disobedient == "1" & disobedienthowmuch == "3","3",  
                                                       ifelse(disobedient == "1" & disobedienthowmuch == "4","4", "check"))))))      

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbCrying_3M = ifelse(crying == "0", "0",  
                                ifelse(crying == "1" & cryinghowmuch == "1", "1",  
                                       ifelse(crying == "1" & cryinghowmuch == "2","2",  
                                              ifelse(crying == "1" & cryinghowmuch == "3","3",  
                                                     ifelse(crying == "1" & cryinghowmuch == "4","4", "check"))))))      

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbWorrySoc_3M = ifelse(worriedsocially == "0", "0",  
                                  ifelse(worriedsocially == "1" & worriedsociallyhowmuch == "1", "1",  
                                         ifelse(worriedsocially == "1" & worriedsociallyhowmuch == "2","2",  
                                                ifelse(worriedsocially == "1" & worriedsociallyhowmuch == "3","3",  
                                                       ifelse(worriedsocially == "1" & worriedsociallyhowmuch == "4","4", "check"))))))      

All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbSep_3M = ifelse(separationworry == "0", "0",  
                             ifelse(separationworry == "1" & separationworryhowmuch == "1", "1",  
                                    ifelse(separationworry == "1" & separationworryhowmuch == "2","2",  
                                           ifelse(separationworry == "1" & separationworryhowmuch == "3","3",  
                                                  ifelse(separationworry == "1" & separationworryhowmuch == "4","4", "check"))))))      


All_Site_Split1 <- All_Site_Split1 %>%  
  mutate(ProbExcFear_3M = ifelse(fearful == "0", "0",  
                                 ifelse(fearful == "1" & fearfulhowmuch == "1", "1",  
                                        ifelse(fearful == "1" & fearfulhowmuch == "2","2",  
                                               ifelse(fearful == "1" & fearfulhowmuch == "3","3",  
                                                      ifelse(fearful == "1" & fearfulhowmuch == "4","4", "check"))))))

All_Site_Split1 <- All_Site_Split1 %>% filter(years_2 >= 5)

#selecting the items for the factor to be analyzed using EFA:
domain_data <- All_Site_Split1 %>% select(ProbHyper_3M, ProbOnTask_3M,ProbAngry_3M,ProbVerbAgg_3M,ProbPhysAgg_3M,ProbDibInjS_3M,ProbDisobArg_3M,ProbCrying_3M, ProbWorrySoc_3M, ProbSep_3M, ProbExcFear_3M)
domain_data<-domain_data[complete.cases(domain_data),]

str(domain_data)
domain_data = domain_data %>% mutate_at(c(1:11), funs(as.numeric(.)))
str(domain_data)
```

```{r}
cor(domain_data)

```

## PARALLEL ANALYSIS ####

```{r message=TRUE, warning=TRUE}
# iterations increased to get a more stable estimate! 
dom_polycor <- hetcor(domain_data, ML=TRUE)
#png(filename=paste0('/Users/cmiguest/Desktop/CRISIS Project/paper_analyses/plots/', analysis_name, '/Parallel_Analysis_results.png'))
dom_par_analysis <- fa.parallel(dom_polycor$correlations, n.obs = nrow(domain_data), n.iter = 1000, error.bars = TRUE)
dev.off()

dom_par_analysis$nfact

#save(dom_par_analysis, file = paste0('../output/', analysis_name, '/parallel_analysis_results.Rdata'))

```




# Test assumptions 
## Bartlett’s test of sphericity

checks whether or not the observed variablesintercorrelate at all using the observed correlation matrix against the identity matrix. If the test found statistically insignificant, you should not employ a factor analysis.
```{r}
# All datasets pass test of significance 
bart_spher(domain_data)
```

## vss Very Simple Structure for number of factors to use
Revelle and Rocklin (1979) applies a goodness of fit test to
determine the optimal number of factors to extract. It can be thought of as a quasiconfirmatory model, in that it fits the very simple structure (all except the biggest loadings per item are set to zero where c is the level of complexity of the item) of a factor pattern matrix to the original correlation matrix. 

These results are confirming that we have a ton of dimensions in our data. dimensions range from 5-8 

```{r, warning = F, message=F}


vss(dom_polycor$correlations, n.obs = nrow(domain_data), rorate ="oblimin")

```


# Factor analysis of the data
https://www.rdocumentation.org/packages/psych/versions/1.8.12/topics/fa
Factor analyze (see section 5.1) the data with a specified number of factors (the
default is 1), the default method is minimum residual, the default rotation for more
than one factor is oblimin.

change to maximum likelihood with fm = "ML"
scores = tenBurgeis the default, we want Bartlett. 


## First decide how many factors
this is using the results from the parallel analysis 
```{r}
NumFactors <- dom_par_analysis$nfact


```

## Run factor analysis with oblimin rotation 
```{r}
# this function requires correlations of data variables
# maximum likelihood 
FA <- psych::fa(r = dom_polycor$correlations, n.obs = nrow(domain_data), nfactors = NumFactors, fm="ML", scores = "Bartlett", rotate = 'oblimin')
FA 

#save(FA, file = paste0('../output/', analysis_name, '/FA_model.Rdata'))

```


## get the eigenvalues 
question is whether this goes in order that we want?
```{r}
# e.values are eigenvalues 
FA_eigen <- data.frame(FA$e.values)
names(FA_eigen) <- "eigenvalues"
# assigning Factor number in numeric order 
FA_eigen$factors <- as.numeric(as.character(rownames(FA_eigen)))

FA_eigen
#save(FA_eigen, file = paste0('../output/', analysis_name, '/FA_eigenvalues.Rdata'))
```

## variance accounted for 
```{r}
FA_variance <- data.frame(FA$Vaccounted)
# changing the names to be in numeric order, based on amount of variance explained.
names(FA_variance) <- c(1:ncol(FA_variance))

# adding labels for the different values provided in this dataset 
FA_variance <- FA_variance %>%
  mutate(value = rownames(data.frame(FA$Vaccounted)))

FA_variance

#save(FA_variance, file = paste0('../output/', analysis_name, '/FA_variance_accounted.Rdata'))
```

## An item by factor (pattern) loading matrix of class loadings. 
note: the factors are numbered BEFORE rotation, so they're actually not in the order we want them to be in! 
```{r}
# show factor loadings with model output 

print(FA, cuttoff = 0, sort = T, digits = 3)
#AFullSample$loadings
```


## get Sample loadings 
```{r}
# get loadings
FA_loadings <- data.frame(matrix(FA$loadings, ncol = NumFactors))
# rename them based on numeric order, of variance explained, not original (un-rotated) labels 
names(FA_loadings) <- c(1:NumFactors)
FA_loadings
# get variable names & weights for later. 
FA_weights <- data.frame(FA$weights)
# rename weights based on numeric order, of variance explained, not original (un-rotated) labels 
names(FA_weights) <- c(1:NumFactors)
FA_weights
FA_weights <- FA_weights %>%
  # pull out variable labels
  mutate(variables = rownames(data.frame(FA$weights))) 

FA_weights
# add thse labels to the loadings too 
FA_loadings$variables <- FA_weights$variables
#save(FA_loadings, file = paste0('../output/', analysis_name, '/FA_loadings_wide.Rdata'))
# convert to long format 
FA_loadings_df_long <- FA_loadings %>%
  gather(key = factor, value = loading, 1:NumFactors)
FA_loadings_df_long

#save(FA_loadings_df_long, file = paste0('../output/', analysis_name, '/FA_loadings_df_long.Rdata'))
```


## get scores for each subjects 
The beta weights to find the factor score estimates. These are also used by the predict.psych function to find predicted factor scores for new cases. These weights will depend upon the scoring method requested.
```{r}

# get raw data into matrix format 
domain_matrix <- domain_data %>%
  as.matrix(.)

## FULL SAMPLE VERSION with labels! 
weights_matrix <- FA_weights %>%
  select(-variables) %>%
    as.matrix(.)

# get pacct labels to add to predictions,

domain_labels <- domain_data %>%
  mutate(subject = as.numeric(rownames(domain_data)))
nrow(domain_labels)

FAsubject_scores <- cbind(domain_labels, domain_matrix%*%weights_matrix)

#save(FAsubject_scores, file = paste0('../output/', analysis_name, '/FA_subject_factor_scores.Rdata'))
                                            
```
